---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = F,
  results = 'hold',
  message = F,
  comment = "",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# RBedtools
<!-- badges: start -->
<!-- badges: end -->
RBedtools is a light wrapper for the [bedtools](https://bedtools.readthedocs.io/en/latest/) software suite. 

I wrote this package for several reasons 

 * I wanted to use bedtools in R. Other packages seemed unnecessarily complicated to use, and native R packages that have simlilar functions as bedtools(read GenomicRanges) were also difficult to use
 * I found myself writing a dataframe to a bed file, running bedtools in a notebook or with system2, then reading it back into R



## Installation

You can install RBedtools from here with:

``` r
# install.packages("devtools")
devtools::install_github("vinay-swamy/RBedtools")
```
Remember to install bedtools as well. Instructions can be found [here](https://bedtools.readthedocs.io/en/latest/content/installation.html)

## Design
I tried to make this package as simple to use as I could. I assumed that the user has some knowledge of how bedtools works and have very little error checking with regards to bedtools input. If you get an error message from bedtools, and not from this package, its likely your bedtools command was wrong. The actual format of bed files/ data frames is not checked, ie 0 based, number of columns etc.

### objects
All RBedtools object are simple S3 objects. 

`bt_obj`: an object that contains a charcter vector that is the path to a bed file.

`bt_cmd`: a object containing a stored and unrun bedtools command. See below for more info

## functions

RBedtools' core function  is `RBedtools` 

```{r args}
library(RBedtools)
args('RBedtools')
```

`tool` is the bedtools function you would like to use, eg intersect, sort, closest etc

`options` are the options for selected tool. It must be formated as a single character vector in the following way:

`'-option1 input1 option2 input2'` the default option is no option

`output` can one of three options: blank, a character vector specifying a location to write a bed file, or `'stdout'` signifying the output of this bedtools command will be piped to another command. More on that below

`...` The actual arguments for the bedtools function, in the following format: `arg1=a.bed, arg2=b.bed`

#### helpers 

there are 2 functions to allow for use of data frame, `from_data_frame`, and `to_data_frame`

`from_data_fram` takes a data frame as an input and returns a bedtools object

`to_data_frame` takes a bedtools object and returns a data frame

## examples

sort a bed file. specifying no output will write to a file in the `/tmp/` directory
```{r basics}
library(RBedtools)
sorted.bed <- RBedtools(tool = 'sort', i='data/gerp.chr1.bed.gz')
unclass(sorted.bed)
```

otherwise the bed is written to the output argument

```{r, eval=F}
sorted.bed <- RBedtools(tool = 'sort',output = 'sorted.bed' i='data/gerp.chr1.bed.gz')
```

use the `to_dataframe` command to convert a `bt_obj` to a dataframe

```{r}
sorted.bed <- RBedtools(tool = 'sort', i='data/gerp.chr1.bed.gz')
sorted.bed.df <- to_data_frame(sorted.bed)
head(sorted.bed.df)
```

piping is also supported(and encouraged)

```{r}
suppressMessages(library(dplyr))
sorted.bed.df <-  RBedtools(tool = 'sort', i='data/gerp.chr1.bed.gz') %>% to_data_frame
head(sorted.bed.df)
```

Use a single character vector for `options` to specfify options for bedtools command. format `options` as you would on the command line

```{r}
intersect_loj <-  RBedtools(tool = 'intersect',options = '-loj', a='data/gerp.chr1.bed.gz', b='data/aluY.chr1.bed.gz') %>% 
    to_data_frame
head(intersect_loj)
```


Using an R data frame

```{r}
closest <-  sorted.bed.df %>% from_data_frame %>% 
    RBedtools('closest', a=., b='data/gerp.chr1.bed.gz') %>% to_data_frame
head(closest)

```

An `RBedtools` command can also be piped to another `RBedtools` command, using `'stdout'` as the output of the first command. When `'stdout'` using `.` place holder for one of the arguments in the second command.

```{r}
sort_and_intersect <-RBedtools(tool = 'sort', output = 'stdout', i='data/gerp.chr1.bed.gz') %>% 
    RBedtools(tool = 'intersect', a=., b='data/aluY.chr1.bed.gz') %>% 
    to_data_frame 
head(sort_and_intersect)
```

An importnant note about piping multiple RBedtools commands

Using `'stdout'` makes `RBedtools` return a `bt_cmd` object, which is a contructed but unrun command 

```{r}
output_for_pipe <-  RBedtools(tool = 'sort', output = 'stdout', i='data/gerp.chr1.bed.gz')
class(output_for_pipe)
```

Commands are not run until the RBedtools commands is run. This allows for the contruction of a single command that uses shell pipes `|` 

```{r}
unclass(output_for_pipe)
```
if you do not specify `'stdout'` when chaining commands, the code will still run, but the output of each command will be written and read each time, which may significantly slow down your code

Full Disclaimer, I havent tested this on all the bedtools functions. I use the sorted, intersect and closest the most, so these are the ones I tested on. If you find a bug/have any improvements/want to let me into your graduate program/ feel free to let me know. 







